Calculating Module Enrichment and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}

### 2016-01-09 jss/article.tex
[skip ci] update examples.

%navicell$launchBrowser()
\section[Examples]{Examples}

@@

The dataset corresponding to this study is available as a Bioconductor package.
The code shown below is creating a list of differentially expressed genes

@@

\begin{tabular}{l|c|c|c|c}
\hline
Module & Mod. size & Nb genes & pval & Cor. pval\\
\hline
Apoptosis:HEDGEHOG & 31 & 21 & 0.0016 & 0.042 \\
Apoptosis:AKT\_MTOR &  79 & 47 & 0.00037 & 0.038 \\
DNA\_repair & 335 & 163 & 0.0012 & 0.038 \\
DNA\_repair:G1\_CC\_PHASE &  25 & 18 & 0.0011 & 0.038 \\
ACSN\_master:AKT\_MTOR & 92 & 53 & 0.0005 & 0.038 \\
ACSN\_master:G1\_CC\_PHASE & 25 & 18 & 0.0011 & 0.038 \\
Survival:S\_PHASE\_CHECKPOINT & 8 & 8 & 0.0006 & 0.038 \\
\hline

@@

>>>>

\section[Examples]{Examples}

@@

The dataset corresponding to this study is available as a Bioconductor package.
The code shown below is creating a list of differentially expressed genes
between ER positive and ER negative samples, and calculates the enrichment in
ACSN maps from this list of genes. As seen in Table \ref{tab:table_mainz}, there
are four maps (Apoptosis, DNA repair, ACSN master and Survival) and seven
modules enriched. Most of the modules are in fact linked to different phases of
the cell cycle.

@@

\begin{tabular}{l|c|c|c|c}
\hline
Module & Mod. size & Nb genes & pval & Cor. pval\\
\hline
Apoptosis:HEDGEHOG & 31 & 21 & 0.0016 & 0.042 \\
Apoptosis:AKT\_MTOR &  79 & 47 & 0.00037 & 0.038 \\
DNA\_repair & 335 & 163 & 0.0012 & 0.038 \\
DNA\_repair:G1\_CC\_PHASE &  25 & 18 & 0.0011 & 0.038 \\
ACSN\_master:AKT\_MTOR & 92 & 53 & 0.0005 & 0.038 \\
ACSN\_master:G1\_CC\_PHASE & 25 & 18 & 0.0011 & 0.038 \\
Survival:S\_PHASE\_CHECKPOINT & 8 & 8 & 0.0006 & 0.038 \\
\hline
\end{tabular}
\end{table}

The Molecular Signatures Database (MSigDB) is one of the most widely used
repository of well-annotated  gene sets representing the universe of biological
processes \citep{liberzon2011molecular}. We downloaded the canonical pathways
set, counting more than 1,300 gene sets representing canonical pathways compiled
by domain experts. The dataset is encoded with the 'gmt' format, and can be
imported within ACSNMineR with the \code{format\_from\_gmt} function. We
calculate the enrichment for the breast cancer differentially expressed gene
list, simply specifying the MSigDB data we just imported as the \code{maps}
option. Table \ref{tab:table_msigdb} is displaying the pathways having a
corrected p-value $< 0.05$. The prefix is indicating the database source, so we
see that we have pathways from the KEGG, Reactome and PID databases. Consistent
with our previous results, most of the enriched pathways are related to the cell
cycle regulation.

@@

\begin{table}[h!]
  \centering
  \caption{MSigDB canonical pathway database enrichment for genes differentially expressed between ER
positive and ER negative samples in breast cancer.  Module : name of the
module. Mod. size: size of the module. Nb genes: number of genes from input
which are found in the module. Cor. pval: corrected p-value.}
\label{tab:table_msigdb}

\begin{tabular}{l|c|c|c}
\hline
Pathway & Mod. size & Nb genes & Cor. pval\\
\hline
KEGG\_CELL\_CYCLE & 128 & 76 & 1.98e-07\\
REACTOME\_CELL\_CYCLE\_MITOTIC & 325 & 158 & 2.40e-07\\
REACTOME\_DNA\_REPLICATION & 192 & 98 & 1.33e-05\\
PID\_FOXM1\_PATHWAY & 40 & 29 & 5.05e-05\\
REACTOME\_MITOTIC\_M\_M\_G1\_PHASES & 172 & 87 & 7.38e-05\\
REACTOME\_CELL\_CYCLE & 421 & 181 & 2.97e-04\\
PID\_AURORA\_B\_PATHWAY & 39 & 27 & 3.38e-04\\
REACTOME\_MITOTIC\_G1\_G1\_S\_PHASES & 137 & 70 & 3.83e-04\\ 
PID\_SYNDECAN\_1\_PATHWAY & 46 & 30 & 4.40e-04\\
REACTOME\_S\_PHASE & 109 & 58 & 4.40e-04\\
\hline
\end{tabular}
\end{table}
